---
title: "Party and Candidate Votes"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article offers a detailed analysis of the [`party_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/party_votes.html) and [`candidate_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/candidate_votes.html) datasets in the `scgElectionsNZ` package, shedding light on the voting dynamics in New Zealand's elections.
<br>
<br>

#### Overview
The [`party_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/party_votes.html) and [`candidate_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/candidate_votes.html) datasets provide comprehensive insights into party preferences and candidate popularity across electorates. These datasets are pivotal for understanding how New Zealand's Mixed-Member Proportional (MMP) system translates individual and party preferences into electoral outcomes.
```{r setup, include = FALSE}
library(scgElectionsNZ)
library(dplyr)
library(ggplot2)
suppressPackageStartupMessages(library(ggpmisc))
```
<br>

#### Loading the Data
Begin by loading the [`party_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/party_votes.html) and [`candidate_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/candidate_votes.html) datasets to understand their structure and the information they contain:

***Party Votes***
```{r party-votes, eval = FALSE}
# Load data
party_df <- scgUtils::get_data("party_votes")

# View data
head(party_df, 10)
```
```{r party-votes_table, echo = FALSE}
party_df <- scgUtils::get_data("party_votes")
knitr::kable(head(party_df, 10))
```

***Candidate Votes***
```{r candidate-votes, eval = FALSE}
# Load data
candidate_df <- scgUtils::get_data("candidate_votes")

# View data
head(candidate_df, 10)
```
```{r candidate-votes_table, echo = FALSE}
candidate_df <- scgUtils::get_data("candidate_votes")
knitr::kable(head(candidate_df, 10))
```
As you can see, the [`candidate_votes`](https://sarahcgall.github.io/scgElectionsNZ/reference/candidate_votes.html) dataset places all unsuccessful parties into "Other" and does not include Informal votes.
<br>
<br>

#### Combining Party and Candidate Vote Data
***Amend Data***

To resolve this issue, use the [`update_names()`](https://sarahcgall.github.io/scgElectionsNZ/reference/update_names.html) function to amend the party names:
```{r party-candidate-votes, eval = FALSE}
# Amend Party names
party_df <- update_names(party_df) # default output = "party"

party_df <- party_df %>%
  filter(Party != "Informal") %>% # Remove "Informals"
  # Get grouped sum of Votes (i.e., combine all "Other" votes into one):
  group_by(Election, Electorate, Ballot, Party) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  # Get grouped percentage and round to 2 decimal places:
  group_by(Election, Electorate, Ballot) %>%
  mutate(Percentage = Votes / sum(Votes) * 100) %>%
  ungroup()

# View amended data
head(party_df, 10)


```
```{r party-candidate-votes_table, echo = FALSE}
party_df <- update_names(party_df)
party_df <- party_df %>%
  filter(Party != "Informal") %>%
  group_by(Election, Electorate, Ballot, Party) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Election, Electorate, Ballot) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  arrange(-Election, Electorate, Ballot, Party)

knitr::kable(head(party_df, 10))
```
***Merge Data***

Then, combine both datasets for a unified analysis:
```{r party-candidate-votes-combined}
# Combine the two data frames
combined_df <- rbind(party_df, candidate_df)
```
<br>

#### Reshaping the Data
Reshape the data to understand how party and candidate preferences align or diverge:
```{r analyse_votes, eval = FALSE}
combined_df <- combined_df %>%
  filter(Election == 2023) %>% # Get the 2023 election only
  # NB Ballot is amended to ensure that there are not two "Party" columns when pivoting wider:
  mutate(Ballot = ifelse(Ballot == "Party", "Party Vote", "Candidate Vote")) %>%
  select(Election, Ballot, Electorate, Party, Percentage) %>% # Remove Votes column
  tidyr::pivot_wider(names_from = Ballot, values_from = Percentage) # Pivot wider

# Make NAs = 0
combined_df[is.na(combined_df)] <- 0

# View data
head(combined_df, 10)
```
```{r analyse_votes_Table, echo = FALSE}
combined_df <- combined_df %>%
  filter(Election == 2023) %>%
  mutate(Ballot = ifelse(Ballot == "Party", "Party Vote", "Candidate Vote")) %>%
  select(Election, Ballot, Electorate, Party, Percentage) %>%
  tidyr::pivot_wider(names_from = Ballot, values_from = Percentage)

combined_df[is.na(combined_df)] <- 0

knitr::kable(head(combined_df, 10))
```
<br>

#### Visualising Party and Candidate Votes
Create a visualisation comparing Party and Candidate votes to understand the correlation between these two types of votes, which can indicate the degree of [split voting](https://sarahcgall.github.io/scgElectionsNZ/articles/split-voting.html) by electorate. The [`colour_pal()`](https://sarahcgall.github.io/scgUtils/reference/colour_pal.html) function from the `scgUtils` package can be used to assign the party colours and the [`theme_scg()`](https://sarahcgall.github.io/scgUtils/reference/theme_scg.html) function from the same package can be used to customise the plot's appearance:
```{r party-candidate-votes1, out.width = '100%', dpi = 300, fig.height = 6}
combined_df %>%
  filter(Party != "Other") %>% # Remove "Other"
  ggplot(aes(x = `Candidate Vote`, y = `Party Vote`, colour = Party)) +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, # Add linear regression
              colour = scgUtils::colour_pal("French Grey"),
              linewidth = 0.5, fill = "#F4F4F5", alpha = 0.5) +
  geom_point(na.rm = TRUE) + # Add scatter plot
  ggpmisc::stat_poly_eq(colour = scgUtils::colour_pal("Regent Grey")) + # Add R^2
  facet_wrap(. ~ Party) + # Facet by Party
  scale_colour_manual(values = scgUtils::colour_pal("polNZ")) + # Get party colours
  coord_equal(ylim = c(0, 80), xlim = c(0, 80)) + # Make axes equal
  labs(title = "2023 Candidate vs Party Vote\nby Electorate and Party", # Add labels
       y = "Party Vote (%)\n",
       x = "Candidate Vote (%)") +
  scgUtils::theme_scg() + # Apply theme
  theme(legend.position = "none") # Turn off legend
```

