---
title: "Seat Allocation Explained"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
```{r setup, include = FALSE}
library(scgElectionsNZ)
library(dplyr)
```
To determine the order in which seats are allocated to parties, the [NZ Electoral Commission](https://electionresults.govt.nz/electionresults_2023/statistics/sainte-lague-formula.html) follows the below process.
<br>
<br>

#### Step 1: Tally Party Votes
First, all Party Votes are tallied for each party and the number of electorate seats won are also tallied.
```{r step1}
# load dataset
df <- scgUtils::get_data("summary")
df <- df %>%
  # filter out total ballot type, leaving the candidate and party ballor
  # types only
  filter(Ballot != "Total",
         # filter for the most recent election
         Election == 2023) %>%
  # keep party votes only by removing candidate votes
  mutate(Votes = ifelse(Ballot == "Candidate", 0, Votes),
         # NB % is not required in this step and is only included
         # for illustrative purposes
         Percentage = ifelse(Ballot == "Candidate", 0, Percentage),
         # keep the number of electorate seats only by removing
         # party/list seats
         Seats = ifelse(Ballot == "Party", 0, Seats)) %>%
  # group long format data frame (NB keep 'Successful' column)
  group_by(Party, Successful) %>%
  summarise(Votes = sum(Votes), Percentage = sum(Percentage),
            Seats = sum(Seats), .groups = 'drop') %>%
  ungroup() %>%
  arrange(-Votes)

head(df)
```
<br>

#### Step 2: Determine Seat Entitlement
Next, ineligible parties (those which do not achieve above 5% Party Vote or win at least 1 electorate) are removed from the denominator of the total votes. The percentage of all party votes is recalculated and the seat entitlement and number of top-up seats are determined.
```{r step2}
df <- df %>%
  # filer out ineligible parties
  filter(Successful == "Yes") %>%
        # New percentage
  mutate(Percentage = Votes/sum(Votes)*100,
         # Number of seats each party is entitled to
         Entitlement = round(120 * Percentage/100, digits=0),
         # Number of top-up seats
         Top_Up = Entitlement - Seats)
head(df)
```
<br>

#### Step 3: Use Saint-Lagu&#235; Formula
Next, the Sainte-Lagu&#235; formula, $Quotient = \frac{Votes}{(2s + 1)}$, is used to determine the order of seats.

NB $s$ begins at 0 for each Party and increases by 1 as seats are allocated each round. This means that the total number of $Votes$ for each party is divided by an odd number every round.

i.e.,

1. $(2 \times 0 + 1) = 1$
2. $(2 \times 1 + 1) = 3$
3. $(2 \times 2 + 1) = 5$

and so on...
```{r step3}
n <- 120 # Total number of seats:
df1 <- data.frame()
# s begins at 0 and sequences up until total number of seats
# has been reached
for(s in seq(0,n)) {
    tmp <- df %>%
      # Saint-Lague Formula:
      mutate(Quotient = Votes / (2*s + 1),
             # included for illustrative purposes only
             s = s) %>%
      select(Party, Quotient, s)
    df1 <- rbind(df1, tmp)
}
head(df1,10)
```
<br>

#### Step 4: Determine Allocation Order
Each seat is then allocated to a party based on the order of the $Quotient$ from highest to lowest. This continues until all 120 seats are allocated (NB this disregards any overhang seats given that these come from excess electorate seats and not list seats).
```{r step4}
# Order seats by Quotient (highest to lowest) and subset
# rows by the total number of seats (120):
df1 <- df1[order(-df1$Quotient),]
df1 <- df1[1:n,]

# Allocate seat number:
df1$Allocation <- 1:n
head(df1,10)
```
<br>

#### Step 5: Allocate Seats
Finally, the electorate seats won by parties are allocated first and the remaining entitlement are allocated from the remaining candidates on party lists (those which did not win an electorate).
```{r step5}

```