---
title: "Split Voting"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article explores the [`split_total`](https://sarahcgall.github.io/scgElectionsNZ/reference/split_total.html) and [`split_electorate`](https://sarahcgall.github.io/scgElectionsNZ/reference/split_electorate.html) datasets.
<br>

```{r setup, include = FALSE}
library(scgElectionsNZ)
library(dplyr)
```
### Split Voting
The `split_total` and `split_electorate` datasets contains the number of voters who cast their party vote for a different party to their candidate vote at the national level, by party. This dataset only has election years 2005 through to 2023 due to there not being split-ticket voting data available for previous years.
<br>
<br>

#### Upload data
View the shape of data frame.
```{r split-voting}
df <- get_data("split_total")
# or
# df <- get_data("split_electorate")
```
<br>

#### Prepare Sankey
Prepare sankey data.
```{r split-voting1}
df <- df %>%
  filter(Election==2023) %>% # get 2023 election data only
  # for electorate-level dataset:
  # filter(Election==2023, Electorate=="Auckland Central") %>%
  # combine unsuccessful minor parties into "Other" category
  mutate(List_Party = ifelse(
    List_Party %in% c("Labour Party","ACT Party","Maori Party",
                      "Green Party","National Party","NZ First",
                      "Informal"),
    List_Party, "Other"
  )) %>%
  mutate(Electorate_Party = ifelse(
    Electorate_Party %in%
      c("Labour Party","ACT Party","Maori Party","Green Party",
        "National Party","NZ First","Informal"),
    Electorate_Party, "Other" # NB includes "Party Vote Only"
  )) %>%
  group_by(List_Party, Electorate_Party) %>%
  summarise(Vote = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  # Percent of party vote going to candidate vote
  group_by(List_Party) %>%
  mutate(Percent = paste0(round(Vote/sum(Vote)*100,2), "%")) %>%
  ungroup()

head(df)
```
<br>

#### View colours
Look up colours using the `scgUtils` package and manually enter.
```{r split-voting2, eval = FALSE}
scgUtils::colour_pal("polNZ")
#> $`ACT Party`
#> [1] "#ffd006"
#>
#> $`Green Party`
#> [1] "#45ba52"
#>
#> $Informal
#> [1] "#d5cdb9"
#>
#> $`Labour Party`
#> [1] "#D82A20"
#>
#> $`Maori Party`
#> [1] "#B2001A"
#>
#> $`National Party`
#> [1] "#00529F"
#>
#> $`NZ First`
#> [1] "#000000"
#>
#> $Alliance
#> [1] "#158f1a"
#>
#> $MANA
#> [1] "#ff0000"
#>
#> $`Jim Anderton's Progressive`
#> [1] "#a91d26"
#>
#> $`United Future`
#> [1] "#4b104e"
#>
#> $Other
#> [1] "#cdcdd1"
```
<br>

#### Plot
Again, using the `scgUtils` package, create sankey to visualise the split vote from the candidate vote (left) to the party vote (right).
```{r split-voting3, eval = FALSE}
scgUtils::plot_sankey(
  data = df,
  source = "Electorate_Party", # left side of sankey
  target = "List_Party", # right side of sankey
  value = "Vote",
  colours = '"#ffd006","#45ba52","#d5cdb9","#D82A20","#B2001A","#000000","#00529F","#cdcdd1"',
  fontSize = 20, # reduce font size from default
  width = 1600 # increase width from default
) # %>%
  # save from viewer to html
  # htmlwidgets::saveWidget(file="sankey_2023.html", selfcontained = TRUE)
```
````{=html}
<iframe width="100%" height="850"
  src="img/sankey_2023.html"
  frameborder="0"
  allowfullscreen></iframe>
````
