---
title: "Split Voting"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article explores the [`split_total`](https://sarahcgall.github.io/scgElectionsNZ/reference/split_total.html) and [`split_electorate`](https://sarahcgall.github.io/scgElectionsNZ/reference/split_electorate.html) datasets.
<br>

```{r setup, include = FALSE}
library(scgElectionsNZ)
library(dplyr)
```
### Split Voting
The `split_total` and `split_electorate` datasets contains the number of voters who cast their party vote for a different party to their candidate vote at the national level, by party. This dataset only has election years 2005 through to 2023 due to there not being split-ticket voting data available for previous years.
<br>
<br>

#### Upload data
View the shape of data frame.
```{r split-voting}
df <- scgUtils::get_data("split_total")
# or
# df <- scgUtils::get_data("split_electorate")
```
<br>

#### Prepare Sankey
Prepare sankey data.
```{r split-voting1}
# Filter 2023 election data only
df <- df[df$Election == 2023, ]

# If electoral-level dataset is used, also filter by Electorate
# df <- df[df$Election == 2023 & df$Electorate == "Auckland Central", ]

# Rename unsuccessful minor parties into "Other" category
df <- update_names(df, nameCol = "List_Party")
df <- update_names(df, nameCol = "Electorate_Party")

df <- df %>%
  # Get sum of votes by flow
  group_by(List_Party, Electorate_Party) %>%
  summarise(Vote = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  # Get percent of party vote going to candidate vote
  group_by(List_Party) %>%
  mutate(Percent = paste0(round(Vote/sum(Vote)*100,2), "%")) %>%
  ungroup()

head(df)
```
<br>

#### Plot
Using the `scgUtils` package, create sankey to visualise the split vote from the candidate vote (left) to the party vote (right). Use the `colour_prep` function from the same package to match the party colours to the correct political party.
```{r split-voting3, eval = FALSE}
scgUtils::plot_sankey(
  data = df,
  source = "List_Party", # left side of sankey
  target = "Electorate_Party", # right side of sankey
  value = "Vote",
  colours = scgUtils::colour_prep(df, c("List_Party","Electorate_Party"), pal_name = "polNZ"),
  width = 900, # reduce width from default of 1200
  height = 700, # reduce height from default of 800
  shiftLabel = 8 # manually shift labels to outside of plot if odd number of parties on each side
) # %>%
  # save from viewer to html
  # htmlwidgets::saveWidget(file="sankey_2023.html", selfcontained = TRUE)
```
````{=html}
<iframe width="100%" height="720"
  src="img/sankey_2023.html"
  frameborder="0"
  allowfullscreen></iframe>
````
NB. Percentage and votes within tooltip on nodes do not exactly equal the `party_vote` or `candidate_vote` of each party due to rounding.
