---
title: "Summary"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article explores the [`summary`](https://sarahcgall.github.io/scgElectionsNZ/reference/summary.html) and [`preMMP_results`](https://sarahcgall.github.io/scgElectionsNZ/reference/preMMP_results.html) datasets.
<br>

```{r setup, include = FALSE}
library(scgElectionsNZ)
library(dplyr)
library(ggplot2)
```

### Summary
The [`summary`](https://sarahcgall.github.io/scgElectionsNZ/reference/summary.html) dataset consolidates key electoral information, providing a bird's-eye view of New Zealand's elections. It can be used to find the number of seats, votes (n & %), and nominations/on party list by ballot type - Candidate Vote, Party Vote, or Total (Seats column = Candidate + Party; Vote column = Party only). It can also be used to find the number of overhang seats and which party these can be attributed to.
<br>
<br>

#### Loading the Data
Begin by loading the [`summary`](https://sarahcgall.github.io/scgElectionsNZ/reference/summary.html) dataset to understand the structure:
```{r summary, eval = FALSE}
# Load datasets by using the following helper function:
df <- scgUtils::get_data("summary")
# Alternatively, use: data("summary"), then df <- summary

# View the data
head(df)
```
```{r summary_table, echo = FALSE}
df <- scgUtils::get_data("summary")
knitr::kable(head(df))
```
<br>

#### Party and Candidate Votes
To view the top performing parties, rank the Party and Candidate Vote by Party for the 2023 election by removing the `Total` ballot type from the dataset:
```{r summary1, eval = FALSE}
df %>%
  filter(Ballot != "Total", Election == 2023) %>%
  arrange(-Percentage) %>%
  head(n = 10)
```
```{r summary1_table, echo = FALSE}
p_c <- df %>%
  filter(Ballot != "Total", Election == 2023) %>%
  arrange(-Percentage) %>%
  head(n = 10)
knitr::kable(p_c)
```
<br>

#### Overhang Seats
As a reminder, overhang seats occur when the Party Vote entitles that party to fewer seats than the number of electorate seats that it won. To view any overhang seats, remove the `Candidate` and `Party` ballot types and unsuccessful parties from the dataset. Then, calculate the overhang seats by multiplying the 120 total parliamentary seats by the proportion of the party vote for successful parties:
```{r summary2, eval = FALSE}
df %>%
  filter(Successful == "Yes", Ballot == "Total", Election == 2023) %>%
  group_by(Election) %>%
  mutate(`Successful_%` = Votes / sum(Votes) * 100) %>%
  ungroup() %>%
  mutate(Overhang = Seats - round(120 * `Successful_%` / 100, 0)) %>%
  arrange(-Overhang)
```
```{r summary2_table, echo = FALSE}
overhang <- df %>%
  filter(Successful == "Yes", Ballot == "Total", Election == 2023) %>%
  group_by(Election) %>%
  mutate(`Successful_%` = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  mutate(Overhang = Seats - round(120 * `Successful_%` / 100, 0)) %>%
  arrange(-Overhang)

knitr::kable(overhang)
```
<br>

***Create Plot***<br>
To view overhang seats over time, visualise the number of seats won by party at each election:
```{r summary3, out.width = '100%', dpi = 300, fig.height = 7}
df %>%
  # remove parties which did not win a seat
  filter(Successful == "Yes", Ballot == "Total") %>%
  ggplot(aes(x = Election, y = Seats,
             fill = factor(Party, levels = c("National Party", "ACT Party", "NZ First",
                                             "United Future", "Jim Anderton's Progressive",
                                             "MANA", "Maori Party", "Alliance", "Green Party",
                                             "Labour Party")))) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.15, alpha = 0.8) +
  # indicate majority required for a 120 seat parliament
  geom_hline(yintercept = 60, colour = "white", linewidth = 0.5) +
  # indicate overhang seats
  geom_hline(yintercept = 120, colour = scgUtils::colour_pal("French Grey")) +
  annotate("text", x = 2025, y = 120, label = "Overhang",
           colour = scgUtils::colour_pal("Regent Grey"),
           size = 3.5, fontface = 2) +
  coord_flip(clip = "off", xlim = c(1994.5, 2024.5)) +
  labs(title = "No. of Seats by Election and Party (1996 - 2023)",
       fill = "Party") +
  scale_fill_manual(values = scgUtils::colour_pal("polNZ")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0), "Election",
                     labels = as.character(df$Election), breaks = df$Election) +
  scgUtils::theme_scg() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```
<br>
<br>

### preMMP Results
The [`summary`](https://sarahcgall.github.io/scgElectionsNZ/reference/summary.html) dataset can be combined with the [`preMMP_results`](https://sarahcgall.github.io/scgElectionsNZ/reference/preMMP_results.html) dataset. The [`preMMP_results`](https://sarahcgall.github.io/scgElectionsNZ/reference/preMMP_results.html) dataset contains the overall results between 1890 and 1993 by party at the national-level.
<br>
<br>

#### Loading the Data
View the shape of data frame.
```{r preMMP, eval = FALSE}
df1 <- scgUtils::get_data("preMMP_results")
head(df1)
```
```{r preMMP_table, echo = FALSE}
df1 <- scgUtils::get_data("preMMP_results")
knitr::kable(head(df1))
```
<br>

#### Merge datasets
To match the `preMMP_results` dataset, remove the `summary` dataset `Party` and `Candidate` ballot types.
```{r preMMP1, eval = FALSE}
df <- df %>%
  filter(Ballot == "Total") %>%
  mutate(Party = ifelse(Successful == "Yes", Party, "Other")) %>%
  group_by(Election, Party) %>%
  summarise(Seats = sum(Seats), Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Election) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  select(Election, Party, Seats, Percentage) %>%
  arrange(-Election)

# Merge
df2 <- rbind(df, df1)

# view dataset
head(df2)
```
```{r preMMP1_table, echo = FALSE}
df <- df %>%
  filter(Ballot == "Total") %>%
  mutate(Party = ifelse(Successful == "Yes", Party, "Other")) %>%
  group_by(Election, Party) %>%
  summarise(Seats = sum(Seats), Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Election) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  select(Election, Party, Seats, Percentage) %>%
  arrange(-Election)
df2 <- rbind(df, df1)

# view dataset
knitr::kable(head(df2))
```
<br>

#### Visualise the Historic Party Vote
Visualise the party/popular vote at each election:
```{r preMMP2, out.width = '100%', dpi = 300, fig.height = 8}
df2 <- df2 %>%
  # filter 1935 until current time (since the beginning of the National and Labour parties)
  filter(Election >= 1935) %>%
  mutate(Party = ifelse(Party %in% c("National Party", "Labour Party"), Party,
                        ifelse(Party == "United-Reform ('National')", "National Party",
                               "Other")),
         Party = factor(Party, levels = c("National Party", "Other", "Labour Party"))) %>%
  group_by(Election, Party) %>%
  summarise(Percentage = round(sum(Percentage), 2), .groups = 'drop') %>%
  ungroup()

df2 %>%
  ggplot(aes(x = reorder(Election, -Election), y = Percentage,
             fill = Party)) +
  geom_bar(stat = "identity", alpha = 0.8, width = 1, size = 0) +
  geom_bar(data = df2 %>% filter(Election == 2023), aes(x = 1, y = Percentage, fill = Party),
           stat = "identity", alpha = 1, width = 1) +
  # Add 50% line
  geom_hline(yintercept = 50, colour = "white") +
  annotate("text", x = 31.2, y = 50, label = "50%", size = 3.5, fontface = 2,
           colour = scgUtils::colour_pal("Regent Grey")) +
  annotate("text", x = 31.2, y = 0, label = "Party Vote >", size = 3, fontface = 2,
           colour = scgUtils::colour_pal("Regent Grey"), hjust = 0) +
  annotate("text", x = 31.2, y = 100, label = "< Party Vote", size = 3, fontface = 2,
           colour = scgUtils::colour_pal("Regent Grey"), hjust = 1) +
  # Add dashed lines for 2023 result
  geom_hline(yintercept = df2$Percentage[df2$Election == 2023 & df2$Party == "Labour Party"],
             colour = "white", linewidth = 0.5, linetype = "dashed") +
  geom_hline(yintercept = 100 - df2$Percentage[df2$Election == 2023 & df2$Party == "National Party"],
             colour = "white", linewidth = 0.5, linetype = "dashed") +
  # Add introduction of MMP line
  geom_vline(xintercept = 10.5, colour = scgUtils::colour_pal("Black80"),
             linetype = "dashed", linewidth = 0.25) +
  annotate("text", x = 10.5, y = 100.75, label = "MMP introduced", size = 3.5, fontface = 2,
           colour = scgUtils::colour_pal("Black80"), hjust = 0) +
  # Add text to 2023 result
  geom_text(x = 1, y = 1,
            label = paste0(df2$Percentage[df2$Election == 2023 & df2$Party == "Labour Party"], "%"),
            hjust = 0, size = 3, colour = "white") +
  annotate("text", x = 0, y = 0, label = "Labour", size = 3.5, fontface = 2,
           colour = "#D82A20", hjust = 0) +
  geom_text(x = 1, y = df2$Percentage[df2$Election == 2023 & df2$Party == "Labour Party"] +
    (df2$Percentage[df2$Election == 2023 & df2$Party == "Other"] / 2),
            label = paste0(format(df2$Percentage[df2$Election == 2023 & df2$Party == "Other"],
                                  nsmall = 2), "%"),
            hjust = 0.5, size = 3, colour = scgUtils::colour_pal("Black80")) +
  annotate("text", x = 0, y = df2$Percentage[df2$Election == 2023 & df2$Party == "Labour Party"] +
    (df2$Percentage[df2$Election == 2023 & df2$Party == "Other"] / 2),
           label = "Minor Parties", size = 3.5, fontface = 2,
           colour = scgUtils::colour_pal("Regent Grey"), hjust = 0.5) +
  geom_text(x = 1, y = 99,
            label = paste0(df2$Percentage[df2$Election == 2023 & df2$Party == "National Party"], "%"),
            hjust = 1, size = 3, colour = "white") +
  annotate("text", x = 0, y = 100,
           label = "National", size = 3.5, fontface = 2,
           colour = "#00529F", hjust = 1) +
  labs(title = "% Party Vote by Election between 1935 and 2023") +
  scale_y_continuous(expand = c(0, 0), position = "right",
                     breaks = seq(0, 100, by = 5)) +
  scale_fill_manual(values = scgUtils::colour_pal("polNZ")) +
  coord_flip(clip = "off", xlim = c(1, 30), ylim = c(0, 100.25)) +
  scgUtils::theme_scg() +
  theme(legend.position = "none",
        plot.margin = unit(c(1, 3, 1, 1), "cm"),
        plot.title = element_text(vjust = 5, colour = scgUtils::colour_pal("Black80")),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank()
  )
```

