---
title: "Vote Type"
---
```{=html}
<style>
  .content p {
    margin-top: 30px !important;
  }

  details {
    margin-bottom: 40px;
  }
</style>
```
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
<br>
This article provides an in-depth exploration of the [`vote_type`](https://sarahcgall.github.io/scgElectionsNZ/reference/vote_type.html) dataset, a key component of the `scgElectionsNZ` package, which delves into the intricacies of voting patterns in New Zealand's elections.
<br>

```{r setup, include = FALSE}
library(scgElectionsNZ)
library(dplyr)
library(ggplot2)
```
### Vote Type
The [`vote_type`](https://sarahcgall.github.io/scgElectionsNZ/reference/vote_type.html) dataset captures the complexities of the New Zealand electoral process by categorising votes based on their validity (Disallowed, Informal, Valid) and method (Ordinary, Special - NZ, Special - Overseas), across electorates, ballot types (Candidate and Party), and elections.
<br>
<br>

#### Loading the Data
Begin by loading the [`vote_type`](https://sarahcgall.github.io/scgElectionsNZ/reference/vote_type.html) dataset and examining its structure:
```{r vote-type, eval = FALSE}
vote_df <- scgUtils::get_data("vote_type")
head(vote_df)
```
```{r vote-type_table, echo = FALSE}
vote_df <- scgUtils::get_data("vote_type")
knitr::kable(head(vote_df))
```
<br>

#### Augmenting Data with Regional Information
Enhance the dataset by adding regional data for more a detailed analysis using the [`add_data()`](https://sarahcgall.github.io/scgElectionsNZ/reference/add_data.html) function:
```{r vote-type3, eval = FALSE}
# Add Regions column
df <- add_data(vote_df, output = "region")

# View Data
head(df)
```
```{r vote-type3_table, echo = FALSE}
df <- add_data(vote_df, output = "region")
knitr::kable(head(df))
```
<br>

#### Visualising Special Votes by Region
Create a visualisation to illustrate the proportion of Party Special Declaration Votes by Region over time.

***Prepare Data***
```{r vote-type4, eval = FALSE}
# Prepare data
df <- df %>%
  filter(Ballot == "Party") %>% # Include Party Votes only
  mutate(Method = ifelse(Method == "Ordinary", Method, "Special")) %>%
  # Get grouped Votes
  group_by(Election, Region, Method) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  # Get grouped Percentage
  group_by(Election, Region) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  filter(Method == "Special") %>% # Get Special Votes only
  arrange(-Percentage)

# View data
head(df)
```
```{r vote-type4_table, echo = FALSE}
df <- df %>%
  filter(Ballot == "Party") %>%
  mutate(Method = ifelse(Method == "Ordinary", Method, "Special")) %>%
  group_by(Election, Region, Method) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Election, Region) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  filter(Method == "Special") %>%
  arrange(-Percentage)

knitr::kable(head(df))
```
***Create Plot***

Use the [`gghighlight()`](https://yutannihilation.github.io/gghighlight/reference/gghighlight.html) function from the `gghighlight` package to highlight single regions in facets and the [`theme_scg()`](https://sarahcgall.github.io/scgUtils/reference/theme_scg.html) function from the `scgUtils` package can be used to customise the plot's appearance:
```{r vote-type_graph, dpi=300, out.width = '100%', fig.height = 7}
df %>%
  ggplot(aes(x = as.character(Election), y = Percentage,
             group = Region, colour = Region)) +
  geom_line() + # Add lines
  geom_point() + # Add points
  gghighlight::gghighlight(use_direct_label = F) + # Highlight by faceted Regions
  facet_wrap(. ~ reorder(Region, -Percentage)) +
  scale_colour_manual(values = c("#374e8e", "#e3b13e", "#df7c18", "#ac004f", # add colours by Region
                                 "#704600", "#1b87aa", "#478c5b"),
                      breaks = c("South Island", "Wellington / Wairarapa",
                                 "Taranaki / Whanganui / Manawatu",
                                 "Bay of Plenty / Hawke's Bay / Gisborne",
                                 "Waikato", "Auckland", "Northland")) +
  labs(title = "Proportion of Special Party Votes\nby Region over Time (1996 - 2023)",
       y = "Percentage (%)\n",
       x = "Election") +
  scgUtils::theme_scg() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5),
        legend.position = "none")
```

As is evident, there are higher proportions of special votes in 1996 and 1999. This is particularly the case for Northland and Taranaki / Whanganui / Manawatu where the number of Maori to General seats is higher. In addition, the urban region of Auckland has a significantly higher proportion of special votes compared with other regions.
<br>
<br>

#### Exploring the Data
To understand the differences in special votes, break down the special votes by method and validity.

**Overseas Voting**<br>
To do this, first examine the proportion of overseas votes by electorate, focusing on the valid and informal votes:
```{r vote-type2, eval = FALSE}
vote_df %>%
  filter(Election == 2023, Validity != "Disallowed") %>% # Only valid and informal votes
  # Get grouped Votes
  group_by(Ballot, Electorate, Method) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  # Get grouped Percentage
  group_by(Ballot, Electorate) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  # View only Special - Overseas results
  filter(Method == "Special - Overseas") %>%
  arrange(-Percentage) %>% # rank by percentage
  head(n = 10)
```
```{r vote-type2_table, echo = FALSE}
overseas <- vote_df %>%
  add_data(output = "region") %>%
  filter(Election == 2023, Validity != "Disallowed") %>%
  group_by(Ballot, Electorate, Region, Method) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Ballot, Electorate) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  filter(Method == "Special - Overseas") %>%
  arrange(-Percentage) %>%
  head(n = 10)
knitr::kable(overseas)
```
The table above shows that the urban electorates within Auckland had the greatest number of overseas voters, likely pushing up the special vote count for the region
<br>
<br>

**Disallowed Votes**<br>
Next, gain insights into the percentage of disallowed votes by electorate and ballot type for the 2023 election:
```{r vote-type1, eval = FALSE}
vote_df %>%
  add_data(output = "type") %>%
  filter(Election == 2023) %>% # Get 2023 Election data only
  # Get grouped Votes
  group_by(Ballot, Electorate, Electorate_Type, Validity) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  # Get grouped Percentage
  group_by(Ballot, Electorate) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  # View only Disallowed results
  filter(Validity == "Disallowed") %>%
  arrange(-Percentage) %>% # rank by percentage
  head(n = 10)
```
```{r vote-type1_table, echo = FALSE}
disallowed <- vote_df %>%
  add_data(output = "type") %>%
  filter(Election == 2023) %>%
  group_by(Ballot, Electorate, Electorate_Type, Validity) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Ballot, Electorate) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  filter(Validity == "Disallowed") %>%
  arrange(-Percentage) %>%
  head(n = 10)

knitr::kable(disallowed)
```
The table above shows the highest proportion of disallowed votes ranked by electorate. Port Waikato shows 100% of the total vote was disallowed. This was due to the ACT Party candidate dying before polling day and thus the poll cancelled for the Candidate Vote (the Party Vote proceeded).

Of note is that the top ten includes all of the M&#257;ori electorates. This is likely due to the higher proportion of M&#257;ori voters casting special votes. Special votes have an added level of complexity and requirements compared with ordinary votes and thus are more likely to be disallowed.
<br>
<br>

#### Visualising Special Votes by Type
And finally, to understand what is driving the higher proportions in 1996 and 1999, add the electorate type, again using the [`add_data()`](https://sarahcgall.github.io/scgElectionsNZ/reference/add_data.html) function:
***Prepare Data***<br>
```{r vote-type6, eval = FALSE}
# Add Electorate Type column
df <- add_data(vote_df, output = "type")

# Prepare data
df <- df %>%
  filter(Ballot == "Party") %>%
  mutate(Method = ifelse(Method == "Ordinary", Method, "Special")) %>%
  group_by(Election, Electorate_Type, Method) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Election, Electorate_Type) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  filter(Method == "Special") %>%
  arrange(-Percentage)

# View data
head(df)
```
```{r vote-type6_table, echo = FALSE}
df <- scgUtils::get_data("vote_type")
df <- add_data(df, output = "type")
df <- df %>%
  filter(Ballot == "Party") %>%
  mutate(Method = ifelse(Method == "Ordinary", Method, "Special")) %>%
  group_by(Election, Electorate_Type, Method) %>%
  summarise(Votes = sum(Votes), .groups = 'drop') %>%
  ungroup() %>%
  group_by(Election, Electorate_Type) %>%
  mutate(Percentage = round(Votes / sum(Votes) * 100, 2)) %>%
  ungroup() %>%
  filter(Method == "Special") %>%
  arrange(-Percentage)

knitr::kable(head(df))
```
***Create Plot***<br>
Visualise the proportion of Party Special Declaration Votes by Electorate Type over time.
```{r vote-type_graph2, dpi=300, out.width = '100%', fig.height = 4}
df %>%
  ggplot(aes(x = as.character(Election), y = Percentage,
             group = Electorate_Type,
             colour = Electorate_Type)) +
  geom_line() +
  geom_point() +
  scale_colour_manual(values = scgUtils::colour_pal("catSimplified")) +
  labs(title = "Proportion of Special Party Votes\nby Electorate Type over Time (1996 - 2023)",
       y = "Percentage (%)\n",
       x = "Election",
       colour = "Electorate Type") +
  scgUtils::theme_scg()
```
This graph shows the significantly higher proportions of special votes cast in M&#257;ori electorates compared with General electorates. The cause of this has been noted in a [parliamentary research paper](https://www.parliament.nz/mi/pb/research-papers/document/00PLLawRP03141/origins-of-the-m%C4%81ori-seats) which explained that due to the geographic size of the M&#257;ori electorates and the limited number of M&#257;ori polling booths, these voters are required to travel further to cast an ordinary vote. As such, M&#257;ori voters were more likely to cast a special vote at General electorate polling places.

Between 1993 and 1999, the number of polling places in which M&#257;ori voters could cast their vote was increased from 534 to 1,203. This was seen to aid in the reduction of special votes.